Rates of true polar wander in convecting planets

### 2014-07-01 tpw_rate.tex
Change to `u`
Add optional detail for Milankovitch theorem

%in case I want to add detail to any derivation
\newif\ifdetail
\detailfalse

@@

\begin{equation}
\nabla \cdot {\bf v} = 0
\label{conserve_mass}
\end{equation}

\begin{equation}
\begin{aligned}
 \rho{\bf \Omega \times \Omega \times r}= - \nabla P + \nabla \cdot \left( \eta \varepsilon ({\bf v}) \right) + \rho {\bf g}
\label{navier_stokes}
\end{aligned}
\end{equation}

\begin{equation}
\frac{\partial T}{\partial t} + {\bf v} \cdot \nabla T = \kappa \nabla^2 T
\label{energy}
\end{equation}

@@

\begin{equation}
 \rho_0{\bf \Omega \times \Omega \times r}= - \nabla P_0 + \rho_0 {\bf g}
\label{navier_stokes}
\end{equation}

@@

Note that the quantity in brackets is identical in form to the shear stress on a plane in classical elastostatics.
If we enter the coordinate system of the convective moment of inertia $\bf E$ with principal moments $\lambda_1 \le \lambda_2 \le \lambda_3$ and define the orientation of $\mitbf{\omega}$ with colatitude $\theta$ and longitude $\phi$ we can write this in a more illuminating form (after some tedious algebra):
\begin{equation}
\begin{aligned}
\dot{\theta} &= - \frac{1}{2 \varepsilon I_0 \tau} \sin{2 \theta} \left[ (\lambda_3-\lambda_1) \cos^2{\phi} + (\lambda_3-\lambda_2) \sin^2{\phi} \right] \\
\dot{\phi} &= \;\; \frac{1}{2 \varepsilon I_0 \tau} \sin{2 \phi} \; (\lambda_2 - \lambda_1)
\end{aligned}
\label{milankovitch}
\end{equation}

>>>>

%in case I want to add detail to any derivation
\newif\ifdetail
\detailfalse
%\detailtrue

@@

\begin{equation}
\nabla \cdot {\bf u} = 0
\label{conserve_mass}
\end{equation}

\begin{equation}
\begin{aligned}
 \rho{\bf \Omega \times \Omega \times r}= - \nabla P + \nabla \cdot \left( \eta \varepsilon ({\bf u}) \right) + \rho {\bf g}
\label{navier_stokes}
\end{aligned}
\end{equation}

\begin{equation}
\frac{\partial T}{\partial t} + {\bf u} \cdot \nabla T = \kappa \nabla^2 T
\label{energy}
\end{equation}

@@

\begin{equation}
 \rho_0{\bf \Omega \times \Omega \times r}= - \nabla P_0 + \rho_0 {\bf g}
\end{equation}

@@

Note that the quantity in brackets is identical in form to the shear stress on a plane in classical elastostatics.
If we enter the coordinate system of the convective moment of inertia $\bf E$ with principal moments $\lambda_1 \le \lambda_2 \le \lambda_3$ and define the orientation of $\mitbf{\omega}$ with colatitude $\theta$ and longitude $\phi$ we can write this in a more illuminating form (after some tedious algebra):

\ifdetail

And here is the tedious algebra.  Let the prefactors $\frac{1}{\varepsilon I_0 \tau}$ be denoted by $A$, and the pole of the coordinate system ($\theta = 0$) associated with the eigenvector for $\lambda_3$.  Therefore we find that $\mitbf{\omega} = \left[ \sin{\theta} \cos{\phi} \;\; \sin{\theta} \sin{\phi} \;\; \cos{\theta} \right]^T$.  Plugging this in, we find a system of three equations, which we want to solve for $\dot{\theta}$ and $\dot{\phi}$:

\begin{equation}
\begin{aligned}
 & \cos{\theta}\cos{\phi} \dot{\theta}  - \sin{\theta}\sin{\phi} \dot{\phi} = \\
  &A \sin{\theta}\cos{\phi}\left[ \lambda_1 - \lambda_1 \sin^2{\theta}\cos^2{\phi} - \lambda_2 \sin^2{\theta}\sin^2{\phi} - \lambda_3 \cos^2{\theta} \right] \\
 &\cos{\theta}\sin{\phi} \dot{\theta}  + \sin{\theta}\cos{\phi} \dot{\phi} = \\
  &A \sin{\theta}\sin{\phi} \left[ \lambda_2 - \lambda_1 \sin^2{\theta}\cos^2{\phi} - \lambda_2 \sin^2{\theta}\sin^2{\phi} - \lambda_3 \cos^2{\theta} \right] \\
 - &\sin{\theta} \dot{\theta} = \\
  &A \cos{\theta} \left[ \lambda_3 - \lambda_1 \sin^2{\theta}\cos^2{\phi} - \lambda_2 \sin^2{\theta}\sin^2{\phi} - \lambda_3 \cos^2{\theta} \right] \\
\end{aligned}
\end{equation}

Now multiply the top equation by $\cos{\phi}$ and the second equation by $\sin{\phi}$ and add them together to get

\begin{equation}
\begin{aligned}
 & \cos{\theta}\cos^2{\phi} \dot{\theta}  - \sin{\theta}\sin{\phi} \cos{\phi} \dot{\phi} + \cos{\theta}\sin^2{\phi} \dot{\theta}  + \sin{\theta}\cos{\phi} \sin{\phi} \dot{\phi} = \\
  &A \sin{\theta}\cos^2{\phi}\left[ \lambda_1 - \lambda_1 \sin^2{\theta}\cos^2{\phi} - \lambda_2 \sin^2{\theta}\sin^2{\phi} - \lambda_3 \cos^2{\theta} \right] + \\
  &A \sin{\theta}\sin^2{\phi} \left[ \lambda_2 - \lambda_1 \sin^2{\theta}\cos^2{\phi} - \lambda_2 \sin^2{\theta}\sin^2{\phi} - \lambda_3 \cos^2{\theta} \right] \\
 & \cos{\theta}\dot{\theta} =  
   A \sin{\theta} \left[ \lambda_1 \cos^2{\phi} +  \lambda_2 \sin^2{\phi} \right] + \\
  &A \sin{\theta} \left[ - \lambda_1 \sin^2{\theta}\cos^2{\phi} - \lambda_2 \sin^2{\theta}\sin^2{\phi} - \lambda_3 \cos^2{\theta} \right] \\
\end{aligned}
\end{equation}

Now multiply the above by $\cos{\theta}$ and the third equation by $-\sin{\theta}$ and add them to get

\begin{equation}
\begin{aligned}
 & \cos^2{\theta}\dot{\theta} + \sin^2{\theta} \dot{\theta} =  
   A \sin{\theta} \cos{\theta} \left[ \lambda_1 \cos^2{\phi} +  \lambda_2 \sin^2{\phi} \right] + \\
  &A \sin{\theta} \cos{\theta} \left[ - \lambda_1 \sin^2{\theta}\cos^2{\phi} - \lambda_2 \sin^2{\theta}\sin^2{\phi} - \lambda_3 \cos^2{\theta} \right] - \\
  &A \cos{\theta} \sin{\theta} \left[ \lambda_3 - \lambda_1 \sin^2{\theta}\cos^2{\phi} - \lambda_2 \sin^2{\theta}\sin^2{\phi} - \lambda_3 \cos^2{\theta} \right] \\
\end{aligned}
\end{equation}

After a lot of cancellation, we find

\begin{equation}
\begin{aligned}
 \dot{\theta} &= A \sin{\theta} \cos{\theta} \left[ - \lambda_3 + \lambda_1 \cos^2{\phi} + \lambda_2 \sin^2{\phi} \right] \\
              &= - A \sin{\theta} \cos{\theta} \left[ (\lambda_3 - \lambda_1) \cos^2{\phi} + (\lambda_3 - \lambda_2) \sin^2{\phi} \right] \\
              &= -\frac{A}{2} \sin{2 \theta} \left[ (\lambda_3 - \lambda_1) \cos^2{\phi} + (\lambda_3 - \lambda_2) \sin^2{\phi} \right] \\
\end{aligned}
\end{equation}

Likewise we may multiply the first equation by $-\sin{\phi}$ and the second equation by $\cos{\phi}$ and add them to get

\begin{equation}
\begin{aligned}
 & \sin{\theta} \dot{\phi} = A \sin{\theta}\cos{\phi}\sin{\phi} \left( \lambda_2 - \lambda_1 \right ) \\
& \dot{\phi} = \frac{A}{2} \sin{2 \phi} (\lambda_2-\lambda_1)
\end{aligned}
\end{equation}

\else
\fi

\begin{equation}
\begin{aligned}
\dot{\theta} &= - \frac{1}{2 \varepsilon I_0 \tau} \sin{2 \theta} \left[ (\lambda_3-\lambda_1) \cos^2{\phi} + (\lambda_3-\lambda_2) \sin^2{\phi} \right] \\
\dot{\phi} &= \;\; \frac{1}{2 \varepsilon I_0 \tau} \sin{2 \phi} \; (\lambda_2 - \lambda_1)
\end{aligned}
\label{milankovitch}
\end{equation}
